Iterative Monomorphisation

### 2025-06-25 iterative_monomorphisation.tex
avoid margin

\begin{sloppy}
\begin{abstract}
Monomorphisation can be used to extend monomorphic provers to support polymorphic logics. We propose a pragmatic iterative approach. We implemented it in the Zipperposition prover, where it is used to translate away polymorphism before invoking the monomorphic prover E as a backend. Our evaluation shows that this approach increases Zipperposition's success rate. Moreover, we find that iterative monomorphisation outperforms native implementations of polymorphism.

@@

We run an example through the detailed version of the iterative algorithm. Consider this initial problem:
\begin{itemize}
   \item[\(\langle1\rangle\)] \(\sym{f}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{f}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
   \item[\(\langle 2\rangle\)] \(\forall x: \alpha{,}\; y:\ty{list}(\alpha){,}\; z: \beta{,}\; \sym{f}\langle \ty{list}(\alpha), \alpha\rangle(y, x) \land \sym{f}\langle \alpha, \alpha \rangle(x, x) \land \sym{f}\langle \alpha, \beta\rangle(x, z)\)
\end{itemize}
%
\(M_{\text{new}}\) is initialised with \(\{f\mapsto (\ty{int}, \ty{nat}); f\mapsto (\ty{int}, \ty{int})\}\) and \(N_{\text{new}}\) with \(\{\langle 2\rangle\mapsto \{f\mapsto (\ty{list}(\alpha), \alpha); f\mapsto (\alpha, \alpha); f\mapsto (\alpha, \beta)\}\}\). Then we enter the main loop (assuming \Loop is at least 1). We iterate over each formula and call \FStep for each of them. Nothing happens for formula \(\langle1\rangle\) because it contains no type variables. For formula \(\langle2\rangle\), \(M_{\text{new}}\) and \(N_{\text{new}}(\langle2\rangle)\) are passed as argument to \SubstGen.

@@

\SubstGen returns and the substitutions are used to generate new type argument tuples. Each substitution is applied to the type arguments of the function symbols of formula \(\langle2\rangle\) because they are the only function symbols with non-monomorphic type arguments.
\begin{table}
\centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{6em}}@{}}
   \toprule
   & \(\sigma_1\) &\(\sigma_2\)&\(\sigma_3\)\\
   \cmidrule(lr){2-2} \cmidrule(lr){3-3} \cmidrule(l){4-4} 
   \((\ty{list}(\alpha), \alpha)\)&\((\ty{list}(\ty{int}), \ty{int})\)& \((\ty{list}(\ty{int}), \ty{int})\) & \((\ty{list}(\ty{int}), \ty{int})\) \\
   \((\alpha, \alpha)\) &\((\ty{int}, \ty{int})\)&\((\ty{int}, \ty{int})\) &\((\ty{int}, \ty{int})\)\\
   \((\alpha, \beta)\)  &\((\ty{int}, \beta)\)&\((\ty{int}, \ty{int})\) &\((\ty{int}, \ty{nat})\)\\
    \bottomrule
\end{tabular}
\end{table}

With a very small and simple initial problem, there are already up to nine new type arguments tuples generated at this step in the first iteration alone, though only four are unique. Once the new type argument tuples are added to their respective maps, a new iteration is begun. We only consider one iteration and continue to the next phase of the algorithm.

\GenFormulae is the next function to be called. This function iterates over all non-monomorphic formulae of the initial problem, in our case this will only be \(\langle2\rangle\). The set of type variable tuples of \(\langle2\rangle\) is passed to \MonoSubst along with the set of all previously generated substitutions.

First, we instantiate \(\alpha\), the first type variable of \(\langle2\rangle\). If \(\sigma_1\) is selected to instantiate \(\alpha\), both \(\sigma_2\) and \(\sigma_3\) will in turn be selected to instantiate \(\beta\). This will generate two monomorphising substitutions: \(\sigma_1 \circ \sigma_2 = \sigma_2\) and \(\sigma_1 \circ \sigma_3 = \sigma_3\) which are added to the set of monomorphising substitutions \(S_{\text{res}}\). Now \(\sigma_2\) is selected. It instantiates both \(\alpha\) and \(\beta\) at once. Here, \(\sigma_2\) is already in \(S_{\text{res}}\) and is therefore ignored, and \(\sigma_3\) is treated similarly. No more than \SubstLimit monomorphising substitutions can be generated in this way.

Our monomorphising substitutions have been generated. We only need to apply them to the formula they monomorphises. This is repeated for all non-monomorphic formulae or until \Limit is reached. The ordering of formulae will impact the output problem in the latter case. For some larger input problems, \MonoSubst could be sufficiently explosive to only allow a handful of different formulae to be monomorphised. The ability to determine \SubstLimit independently can help avoid this issue.

Finally, all monomorphic formulae, both new and old, are mangled. This step is unecessary if the target language allows type constructors. In our example two new monomorphic formulae are generated. The algorithm outputs:
\begin{enumerate}
   \item \(\sym{f}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{f}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
   \item[3.] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{int}{,}\; \sym{f}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{f}\langle \ty{int}, \ty{int} \rangle(x, x) \land \sym{f}\langle \ty{int}, \ty{int}\rangle(x, z)\)
   \item[4.] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{nat}{,}\; \sym{f}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{f}\langle \ty{int}, \ty{int} \rangle(x, x) \land \sym{f}\langle \ty{int}, \ty{nat}\rangle(x, z)\)
\end{enumerate}

@@

\looseness=-1
Table~\ref{nmon_ty_args} plays the same role as Table~\ref{mono_ty_args} except that it evaluates the bounds limiting the number of new \emph{non-}monomorphic type arguments. The bound values are lower because we found non-monomorphic type arguments to be combinatorially explosive in preliminary evaluations. The table confirms that non-monomorphic type argument generation drives the creation of useful non-mono\-morphic formulae. This is indicated by the very low number of problems solved when no new non-monomorphic type arguments are allowed. Performance of the monomorphisation algorithm seems to plateau for some ranges of values and drops off beyond.
%TODO check that 125 is what we get if we just run Zipperposition for 10~s (maybe with all additional problems that are easily monomorphised)

>>>>

\begin{sloppy}
\begin{abstract}
Monomorphisation can be used to extend monomorphic provers to support polymorphic logics. We describe a pragmatic iterative approach. We implemented it in the Zipperposition prover, where it is used to translate away polymorphism before invoking the monomorphic prover E as a backend. Our evaluation shows that this approach increases Zipperposition's success rate. Moreover, we find that iterative monomorphisation outperforms native implementations of polymorphism.

@@

To illustrate the low level version of the iterative algorithm, we consider the following initial problem:
\begin{itemize}
   \item[\(\langle1\rangle\)] \(\sym{p}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{p}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
   \item[\(\langle 2\rangle\)] \(\forall x: \alpha{,}\; y:\ty{list}(\alpha){,}\; z: \beta{,}\; \sym{p}\langle \ty{list}(\alpha), \alpha\rangle(y, x) \land \sym{p}\langle \alpha, \alpha \rangle(x, x) \land \sym{p}\langle \alpha, \beta\rangle(x, z)\)
\end{itemize}
%
\(M_{\text{new}}\) is initialised with \(\{\sym p\mapsto (\ty{int}, \ty{nat}); \sym p\mapsto (\ty{int}, \ty{int})\}\) and \(N_{\text{new}}\) with \(\{\langle 2\rangle\mapsto \{\sym p\mapsto (\ty{list}(\alpha), \alpha); \sym p\mapsto (\alpha, \alpha); \sym p\mapsto (\alpha, \beta)\}\}\). Then we enter the main loop (assuming \Loop is at least 1). We iterate over each formula and call \FStep for each of them. Nothing happens for formula \(\langle1\rangle\) because it contains no type variables. For formula \(\langle2\rangle\), \(M_{\text{new}}\) and \(N_{\text{new}}(\langle2\rangle)\) are passed as argument to \SubstGen.

@@

Then \SubstGen returns, and the substitutions are used to generate new type argument tuples. Each substitution is applied to the type arguments of the function symbols of formula \(\langle2\rangle\) because they are the only function symbols with non-monomorphic type arguments.
The table below summarises the situation:
\begin{table}
\centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{6em}}@{}}
   \toprule
   & \(\sigma_1\) &\(\sigma_2\)&\(\sigma_3\)\\
   \cmidrule(lr){2-2} \cmidrule(lr){3-3} \cmidrule(l){4-4} 
   \((\ty{list}(\alpha), \alpha)\)&\((\ty{list}(\ty{int}), \ty{int})\)& \((\ty{list}(\ty{int}), \ty{int})\) & \((\ty{list}(\ty{int}), \ty{int})\) \\
   \((\alpha, \alpha)\) &\((\ty{int}, \ty{int})\)&\((\ty{int}, \ty{int})\) &\((\ty{int}, \ty{int})\)\\
   \((\alpha, \beta)\)  &\((\ty{int}, \beta)\)&\((\ty{int}, \ty{int})\) &\((\ty{int}, \ty{nat})\)\\
    \bottomrule
\end{tabular}
\end{table}

With a very small and simple initial problem, there are already up to nine new type arguments tuples generated at this step in the first iteration alone, although only four are unique. Once the new type argument tuples are added to their respective maps, a new iteration is begun. We only consider one iteration and continue to the next phase of the algorithm.

The next function to be called is \GenFormulae. It iterates over all non-monomorphic formulae of the initial problem; in our case, this will only be \(\langle2\rangle\). The set of type variable tuples of \(\langle2\rangle\) is passed to \MonoSubst along with the set of all previously generated substitutions.

First, we instantiate \(\alpha\), the first type variable of \(\langle2\rangle\). If \(\sigma_1\) is selected to instantiate \(\alpha\), both \(\sigma_2\) and \(\sigma_3\) will in turn be selected to instantiate \(\beta\). This will generate two monomorphising substitutions, \(\sigma_1 \circ \sigma_2 = \sigma_2\) and \(\sigma_1 \circ \sigma_3 = \sigma_3\), which are added to the set of monomorphising substitutions \(S_{\text{res}}\). Now \(\sigma_2\) is selected. It simultaneously instantiates \(\alpha\) and \(\beta\). Here, \(\sigma_2\) is already in \(S_{\text{res}}\) and is therefore ignored, and \(\sigma_3\) is treated similarly. No more than \SubstLimit monomorphising substitutions can be generated in this way.

The monomorphising substitutions have been generated. We only need to apply them to the formula they monomorphise. This is repeated for all non-monomorphic formulae or until \Limit is reached. The ordering of formulae will impact the output problem in the latter case. For some larger input problems, \MonoSubst could be sufficiently explosive to only allow a handful of different formulae to be monomorphised. The ability to set \SubstLimit independently can help avoid this issue.

Finally, if the target language supports only nullary type construtors, the types of all monomorphic formulae, both new and old, are mangled.
%This step is unecessary if the target language allows type constructors.
In our example, two new monomorphic formulae are generated. The algorithm outputs
\begin{enumerate}
   \item[\(\langle1\rangle\)] \(\sym{p}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{p}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
   \item[\(\langle3\rangle\)] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{int}{,}\; \begin{aligned}[t]
     & \sym{p}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{p}\langle \ty{int}, \ty{int} \rangle(x, x) \land {} \\[-\jot]
     & \sym{p}\langle \ty{int}, \ty{int}\rangle(x, z)\end{aligned}\)
   \item[\(\langle4\rangle\)] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{nat}{,}\; \begin{aligned}[t]
     & \sym{p}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{p}\langle \ty{int}, \ty{int} \rangle(x, x) \land {} \\[-\jot]
     & \sym{p}\langle \ty{int}, \ty{nat}\rangle(x, z)\end{aligned}\)
\end{enumerate}

@@

\looseness=-1
Table~\ref{nmon_ty_args} is similar to Table~\ref{mono_ty_args} except that it evaluates the bounds limiting the number of new \emph{non-}monomorphic type arguments. The bound values are lower because we found non-monomorphic type arguments to be combinatorially explosive in preliminary evaluations. The table confirms that non-monomorphic type argument generation drives the creation of useful non-mono\-morphic formulae. This is indicated by the very low number of problems solved when no new non-monomorphic type arguments are allowed. Performance of the monomorphisation algorithm seems to plateau for some ranges of values and drops off beyond.
%TODO check that 125 is what we get if we just run Zipperposition for 10~s (maybe with all additional problems that are easily monomorphised)
